An Evolutionary Algorithm for the Multiobjective Risk-Equity Constrained Routing Problem

### 2011-11-30 trunk/DAW_hazmatHypE/trunk/paper/hazmatRouting_EvolOptim.tex
finished first version of "feasibility of the approach" section

Instead of the environmental selection of HypE \citep{bz2011a}, other selection schemes can be easily used with the PISA framework and we show the resutls of two other algorithms exemplary. On the one hand, we employ the famous NSGA-II selection scheme \citep{dapm2002a} and, on the other hand, we use a generalized version of HypE towards a generalized weighted hypervolume indicator \citep{abbz2009b,bbtz2011a}. The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations) are shown in Fig.~\ref{fig:algoComparison}. \TODO{What can be seen is that both? algorithms show a reduced performance regarding the standard hypervolume value. For NSGA-II, this is expected as its environmental selection schemes becomes less effective than in bi-objective problems when the number of objectives increases \citep{wbn2007a,bz2011a}. This is not the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing non-compliant quality measures such as NSGA-II.

The reason for the reduced performance in the standard hypervolume indicator for W-Hype, however, is a different one. W-HypE was developed to steer the search towards regions of the objective space that have been specified by the user. To this end, it does not optimize the standard hypervolume indicator but a more generalized version of it, the weighted hypervolume indicator \citep{zbt2007a}. The idea behind W-HypE is that solutions in objective space regions with a higher weight will get a higher fitness due to the \emph{weighted} hypervolume they dominate. In order to be efficient also for a large number of objectives, W-HypE does not calculate the weighted hypervolume indicator values exactly (which is not possible for most of the weight functions) but instead samples them. Here, 10,000 samples are used in each iteration of the algorithm.

To show the principles of W-HypE in the application of our hazmat routing problem, we defined the weight of W-HypE's underlying weighted hypervolume indicator as constant in a box with corners $[7\cdot 10^3, 3\cdot 10^7, 5\cdot 10^5]$ and $[10^4, 4\cdot 10^7, 1.1\cdot 10^6]$ and zero otherwise. Instead of optimizing the standard hypervolume indicator as HypE, W-HypE now optimizes this weighted hypervolume indicator, resulting in better, i.e., dominating solutions in regions in or close to the box. Figure~\ref{fig:whype} shows the resulting non-dominated solutions of both algorithms with population size 25 after 10 independent runs of 4,000 generations each as well as the sampling box.}

This advantage of W-HypE over HypE within the pre-defined box can be observed both in early and late generations, cp.\ Fig.~\ref{fig:whype}.

\TODO{write}

\TODO{W-HypE: number of non-dominated solutions in the sampling box: 14 for W-HypE, 6 for HypE; when overall non-dominated solutions are computed, W-HypE has 13, HypE only 1}

@@

\subsection{The Feasibility of the EMO Approach}
Due to their heuristic nature, evolutionary multiobjective optimization algorithms like the one presented in this study cannot guarantee that the solutions found are actually Pareto-optimal. However, as we have seen above, the proposed algorithm is able to produce a set of non-dominated solutions in reasonable time which are close to Pareto-optimal solutions found by exact algorithms. Such sets of non-dominated solutions can be a starting point for a decision maker to investigate the trade-offs between the objectives and to decide in which objective space region interesting Pareto-optimal solutions are which can be explored further with more involved algorithms---be it exact algorithms or specialized evolutionary algorithms such as the mentioned W-HypE variant. In the above example instances, it turned out that total risk and cost are indeed more or less fully correlated in the original problem formulation or only loosely (with a larger trade-off between cost and total risk) when the individual arcs' risks and costs are.

Regarding the two initialization schemes, it turned out that in combination with the (local and simple) variation operator, the cost-efficient initialization does not allow the algorithm to find additional solutions besides the initial one and that, in particular if the cost and total risk objectives are correlated, the random initialization is beneficial. Using other initializations as well as additional variation operators from the literature is easily possible with the proposed algorithm but their investigation is left for future research\footnote{For example, the

On the other hand, the proposed variation operator can be changed towards a larger neighborhood by allowing two or more nodes to be added or deleted from each truck path at a time---resulting in more exploration than exploitation when compared to the current variation operator. Also well-designed crossover operators are expected to be beneficial to the problem when, for example, good sub paths can be exchanged between solution paris.
}.

\begin{itemize}

	\item approach is very general and can easily deal with slightly modified problems: e.g. different truck loads (only the objective function code needs to be changed)
	\item it needs to be mentioned that for the tested instances with 311 nodes and 441 arcs, the variation operator of the evolutionary algorithm was able to construct the shortest paths from source to destination in some of the runs for some of the truck paths within the first 100,000 function evaluations. This corresponds to performing random walks of lengths around 20 nodes.
\end{itemize}

@@

>>>>

Instead of the environmental selection of HypE \citep{bz2011a}, other selection schemes can be easily used with the PISA framework and we show the resutls of two other algorithms exemplary. On the one hand, we employ the famous NSGA-II selection scheme \citep{dapm2002a} and, on the other hand, we use a generalized version of HypE towards a generalized weighted hypervolume indicator \citep{abbz2009b,bbtz2011a}. The boxplots of the resulting hypervolume indicator values after 100,000 function evaluations (population of size 100 after 1000 generations for NSGA-II and population size of 25 and 4,000 generations for W-HypE) are shown in Fig.~\ref{fig:algoComparison}. What can be seen is that both algorithms show a reduced performance regarding the standard hypervolume value with W-HypE being much worse than NSGA-II. For NSGA-II, this decrease  in hypervolume is expected as its environmental selection schemes does not optimize the hypervolume indicator as does HypE. Moreover, NSGA-II is known to become less effective than in bi-objective problems when the number of objectives increases \citep{wbn2007a,bz2011a}. This is not the case for indicator-based algorithms such as HypE which were specifically proposed to circumvent the problems of algorithms optimizing quality measures that do not comply with the Pareto dominance relation such as the crwoding distance of NSGA-II.

The reason for the reduced performance in the standard hypervolume indicator for W-HypE, on the other hand, can be explained in a different way. W-HypE and the underlying weighted hypervolume indicator was developed to steer the search towards regions of the objective space that have been specified by the user. To this end, instead of the standard hypervolume indicator, the weighted hypervolume indicator of \citep{zbt2007a} is optimized. The idea behind W-HypE is that solutions in objective space regions with a higher weight will get a higher fitness due to the \emph{weighted} hypervolume they dominate. Furthermore, in order to be efficient also for a large number of objectives, W-HypE does not calculate the weighted hypervolume indicator values exactly but instead samples them. Here, 10,000 samples are used in each iteration of the algorithm. For details on the algorithm, we refer to the original publication \citep{bz2011a}.

To show the principles of W-HypE in the application of our hazmat routing problem, we defined the weight of W-HypE's underlying weighted hypervolume indicator as constant in a box with corners $[7\cdot 10^3, 3\cdot 10^7, 5\cdot 10^5]$ and $[10^4, 4\cdot 10^7, 1.1\cdot 10^6]$ and zero otherwise. Instead of optimizing the standard hypervolume indicator as HypE does, W-HypE now optimizes this weighted hypervolume indicator as a secondary selection criterion after employing a non-dominated sorting. As a result, it is finding better, i.e., dominating solutions in regions in or close to the box. Figure~\ref{fig:whype} shows the resulting non-dominated solutions of both algorithms W-HypE and HypE with population size 25 after 10 independent runs of 4,000 generations together with the sampling box. While HypE originally finds 6 non-dominated solutions within the box, W-HypE finds 14 within the same budget of function evaluations. When both algorithm outcomes are combined, it turns out that 13 of W-HypE's solutions are still non-dominated while 5 out of the 6 solutions found by HypE are dominated by W-HypE solutions. It has to be noted that the advantage of W-HypE over HypE within the pre-defined box can be observed both in early and late generations (results not shown here).

@@

\subsection{The Feasibility of the EMO Approach}
Due to their heuristic nature, evolutionary multiobjective optimization algorithms like the one presented in this study cannot guarantee that the solutions found are actually Pareto-optimal. However, as we have seen above, the proposed algorithm is able to produce a set of non-dominated solutions in reasonable time. In particular the comparison with the cost-optimal solution shows that the randomized algorithm can find solutions close to the Pareto front. The found sets of non-dominated solutions can be a starting point for a decision maker to investigate the trade-offs between the objectives and to decide in which objective space region interesting Pareto-optimal solutions should be searched for with more involved algorithms---be it exact algorithms or specialized evolutionary algorithms such as the mentioned W-HypE variant.

Regarding the two initialization schemes, it turned out that in combination with the (local and quite simple) variation operator, the cost-optimal initialization does not allow the algorithm to find additional solutions besides the initial one. However, if the cost and total risk objectives are less correlated, this cost-optimal initialization has advantages over the random initialization in the early stage of the search when compared with respect to the hypervolume indicator. In general, however, we suggest to use the random initialization if the number of function evaluations is not a limiting factor.

Using other initializations as well as additional variation operators from the literature is easily possible with the proposed algorithm but their investigation is left for future research. For example, one can think of a ``more random'' initialization where the initial truck paths are generated by iteratively applying the mutation operator on the empty path and also the multiple application of the described mutation operator within one single mutation might be beneficial as the explored space of neighbored paths is larger in comparison to the current mutation. Also well-designed crossover operators are expected to be beneficial to the problem when, for example, good sub paths can be exchanged between solution pairs.

Overall, it is to say that the approach is very general and can easily be adapted if the problem formulationi is (slightly) changed as it is often the case in practice, e.g., if different numbers of trucks are available. Also in case of different truck loads, it is only the objective function code that needs to be changed---a big advantage of blackbox optimizers such as the proposed EMO algorithm.

@@

\TODO{ In the above example instances, it turned out that total risk and cost are indeed more or less fully correlated in the original problem formulation or only loosely (with a larger trade-off between cost and total risk) when the individual arcs' risks and costs are.
}
